---
title: 'Poor performance of Wald CIs for proportions'
author: "Jacob Goldstein-Greenwood"
date: last-modified
date-format: iso
format:
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

Andersson, 2023, _The American Statistician_, "The Wald confidence interval for a binomial _p_ as an illuminating 'bad' example." DOI: [10.1080/00031305.2023.2183257](https://www.tandfonline.com/doi/full/10.1080/00031305.2023.2183257).

> "The standard two-sided Wald interval for the binomial _p_ is by now well known to be of low quality with respect to actual coverage and it ought to be stricken from all courses in mathematical statistics/statistics as the recommended interval to use. However, it still lingers on in quite a few textbooks."

The Wald confidence interval (CI) for a proportion is erratic, often straying markedly from the nominal coverage rate. Received wisdom is that the interval's performance decays as _p_ drifts from 0.5 toward 0 or 1 and/or as _n_ decreases. Brown et al. (2001, _Statistical Science_) note that it's more accurate to characterize the Wald interval as being subject to "lucky" and "unlucky" (_p_, _n_) pairings. For example, they show that, surprisingly, the 95% Wald interval's coverage is higher (and closer to 95%) for (_p_ = 0.2, _n_ = 30) than for (_p_ = 0.2, _n_ = 98).

I simulated data from binomial distributions at a range of (_p_, _n_) combinations and estimated 95% Wald CIs and 95% Wilson CIs for the resulting proportions. I repeated the process 2500 times for each (_p_, _n_) pairing. The coverage rates---the fraction of intervals of each type that captured _p_ for each set of (_p_, _n_) values---are below:

```{r, warning=FALSE, message=FALSE}
library(ggplot2)

# Set simulation conditions
set.seed(999)
props <- seq(.05, .95, by = .025)
n <- c(25, seq(50, 550, by = 100))
sim_conditions <- expand.grid(props, n)
colnames(sim_conditions) <- c('prop', 'n')
res <- vector('list', nrow(sim_conditions))
reps <- 2500
z_975 <- qnorm(.975)

# Run
#   Formula reference for Wilson CI:
#   https://www.itl.nist.gov/div898/handbook/prc/section2/prc241.htm
for (i in 1:nrow(sim_conditions)) {
  res[[i]] <- replicate(reps, {
    draws <- rbinom(sim_conditions$n[i], 1, sim_conditions$prop[i])
    prop <- mean(draws)
    
    wald_CI_width <- z_975 * sqrt((prop*(1-prop)) / sim_conditions$n[i])
    wald_CI <- prop + c(-wald_CI_width, wald_CI_width)
    wald_captures <- ifelse(sim_conditions$prop[i] <= wald_CI[2] &&
                              sim_conditions$prop[i] >= wald_CI[1], T, F)
    
    wilson_CI_width <- z_975*sqrt((prop*(1-prop))/sim_conditions$n[i] +
                                    z_975^2/(4*sim_conditions$n[i]^2))
    wilson_CI <- (prop + z_975^2/(2*sim_conditions$n[i]) +
                    c(-wilson_CI_width, wilson_CI_width)) / (1 + z_975^2/sim_conditions$n[i])
    wilson_captures <- ifelse(sim_conditions$prop[i] <= wilson_CI[2] &&
                                sim_conditions$prop[i] >= wilson_CI[1], T, F)
    
    c('Wald' = wald_captures, 'Wilson' = wilson_captures)
  })
}

# Organize results
sim_conditions$coverage_Wald <- sapply(res, \(x) mean(x[1, ]))
sim_conditions$coverage_Wilson <- sapply(res, \(x) mean(x[2, ]))

d <- tidyr::pivot_longer(sim_conditions,
                         cols = c('coverage_Wald', 'coverage_Wilson'),
                         values_to = 'coverage',
                         names_to = 'method',
                         names_prefix = 'coverage_')

d$n <- as.factor(d$n)

# Plot
ggplot(d, aes(prop, coverage, color = n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~method) +
  scale_x_continuous(breaks = seq(.05, .95, by = .15)) +
  ylim(.5, 1) +
  labs(x = 'Population proportion',
       y = '95% CI coverage rate',
       title = '95% Wald and Wilson CI coverage rates for proportions',
       color = 'N',
       caption = 'Each point reflects 2500 simulations') +
  theme(legend.position = 'bottom') +
  guides(color = guide_legend(nrow = 1)) +
  geom_hline(yintercept = .95)
```

The simulation results show, as noted by Brown et al. (2001) and Andersson (2023), that the Wald interval's performance is jagged: It does not decrease monotonically as _p_ approaches 0 or 1, nor does it decrease monotonically as _n_ decreases. That one can find occasional "lucky" (_p_, _n_) combinations does not, though, make the picture rosy: The Wald CI has broadly decayed coverage, and many more of the Wald CI coverage rates calculated here fall below 95% compared to the Wilson CI coverage rates:

```{r, warning=FALSE, message=FALSE}
under_coverage_prop <- aggregate(d$coverage,
                                 list(d$method),
                                 \(x) round(mean(x < .95), digits = 3))

knitr::kable(under_coverage_prop,
             col.names = c('CI type', 'Proportion of coverage rates < .95'))
```

The Wilson interval is not alone as a (preferable) alternative to the Wald interval for proportions, but it is a fair go-to. It received complimentary examinations in Andersson (2023) and Brown et al. (2001), and it is Harrell's [recommended (frequentist) interval](https://hbiostat.org/bbr/htest#frequentist-methods) as of April 2025 in _Biostatistics for Biomedical Research_.